---
title: "Exercise with Parallel Coordinate Plots (PCP)"
subtitle: "*An example*"
author: "Christophe Bontemps (SIAP)"
date: last-modified
format:
  revealjs:
    toc: false
    slide-number: true
    theme: default
    incremental: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Parallel Coordinate Plots (PCP) are a powerful visualization technique for exploring multivariate data.

- Each variable is a vertical axis
- Each observation is a connected line across the axes
- Useful for spotting patterns across **many dimensions at once**

<!-- ## Load Libraries -->

<!-- Before running this code, install 3 packages: -->

<!-- - `tidyverse` – main package for data science -->
<!-- - `GGally` – provides `ggparcoord()` for PCPs -->
<!-- - `plotly` – for interactive graphics -->

```{r load-libraries}
library(tidyverse)
library(GGally)      # For ggparcoord
library(plotly)      # For interactive plots
```

## Data

We generate the data for the sake of the exercise.

```{r generate-data}
set.seed(123)

data <- tibble(
  id = 1:10,
  sex = c(rep("Men", 6), rep("Women", 4)),

  # Age: Men older (45-55), Women younger (25-35)
  age = c(48, 52, 45, 50, 47, 54,
          28, 32, 26, 30),

  # Education: Men lower (12-14 yrs), Women higher (16-18 yrs)
  education = c(12, 14, 13, 12, 14, 13,
                18, 17, 18, 16),

  # Income: Men high (70k-90k), Women moderate (40k-55k)
  income = c(85000, 78000, 82000, 90000, 75000, 88000,
             48000, 52000, 45000, 50000)
)

write.csv(select(data, age, education, income),
          "Data/PCPData.csv", row.names = FALSE)
```

## Basic Parallel Coordinate Plot

```{r basic-pcp}
#| fig-align: center
data_pcp <- data %>%
  select(id, age, education, income)

ggparcoord(
  data_pcp,
  columns = 2:4,
  groupColumn = "id",
  scale = "uniminmax"
) +
  labs(
    title = "Age, Education, and Income",
    x = NULL, y = "Scaled Values (0-1)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.notes}
Notice how the lines form two distinct groups, even though we haven't colored them.
:::

## Enhanced PCP with Transparency

Adding transparency (`alpha`) helps reveal overlapping patterns and density.

```{r enhanced-pcp}
#| fig-align: center
ggparcoord(
  data_pcp,
  columns = 2:4,
  groupColumn = "id",
  scale = "uniminmax",
  alphaLines = 0.6
) +
  geom_line(linewidth = 1.2) +
  labs(title = "With transparency", x = NULL, y = "Scaled Values") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Interactive Version {.smaller}

Click the button to reveal the hidden grouping by **sex**.

```{r interactive-pcp}
#| fig-height: 5 
#| fig-width: 8
plot_data <- data %>%
  mutate(
    color_code = ifelse(sex == "Men", 0, 1),
    label = paste0(
      "ID: ", id, "<br>",
      "Sex: ", sex, "<br>",
      "Age: ", age, "<br>",
      "Education: ", education, " years<br>",
      "Income: $", format(income, big.mark = ",")
    )
  )

n <- nrow(plot_data)

fig <- plot_ly(
  data = plot_data,
  type = 'parcoords',
  line = list(
    color = rep(0, n),
    colorscale = list(c(0, '#999999'), c(1, '#999999')),
    showscale = FALSE,
    width = 4
  ),
  dimensions = list(
    list(label = 'Age', values = ~age, range = c(20, 60)),
    list(label = 'Education (years)', values = ~education, range = c(10, 20)),
    list(label = 'Income ($)', values = ~income, range = c(30000, 90000))
  )
)

fig <- fig %>%
  layout(
    title = list(text = "Parallel Coordinate Plot", font = list(size = 16), y = 0.98),
    annotations = list(list(
      text = "Drag axes to filter",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0.8, y = 1.15, font = list(size = 9, color = "grey40")
    )),
    updatemenus = list(list(
      type = "buttons", direction = "right",
      x = 0.5, xanchor = "center", y = 1.3, yanchor = "top",
      buttons = list(
        list(method = "restyle",
             args = list(list("line.color" = list(rep(0, n)),
                               "line.colorscale" = list(list(c(0, '#999999'), c(1, '#999999'))))),
             label = "No Color"),
        list(method = "restyle",
             args = list(list("line.color" = list(plot_data$color_code),
                               "line.colorscale" = list(list(c(0, '#3498DB'), c(1, '#E74C3C'))))),
             label = "Color")
      )
    )),
    margin = list(l = 100, r = 100, t = 110, b = 30),
    height = 380,
    autosize = TRUE,
    plot_bgcolor = '#f8f9fa',
    paper_bgcolor = 'white'
  )

fig
```

## PCP with Hidden Grouping

Coloring the lines by sex makes the two groups visually distinct.

```{r pcp-colored}
#| fig-align: center
data_with_sex <- data %>%
  select(id, sex, age, education, income)

ggparcoord(
  data_with_sex,
  columns = 3:5,
  groupColumn = "sex",
  scale = "uniminmax",
  alphaLines = 0.7
) +
  scale_color_manual(values = c("Men" = "#3498DB", "Women" = "#E74C3C")) +
  labs(title = "Two Distinct Groups", x = NULL, y = "Scaled Values (0-1)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Key Patterns

- **Blue lines** (Men): Higher age and income, moderate education
- **Red lines** (Women): Lower age and income, higher education

## Summary Statistics by Group

```{r summary-stats}
data %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    avg_age = mean(age),
    avg_education = mean(education),
    avg_income = mean(income),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 1,
    col.names = c("Sex", "N", "Avg Age", "Avg Education", "Avg Income")
  )
```

## Key Insights

1. **Two Distinct Patterns**: Even without labeling sex, two clear groups emerge
2. **Age–Income Relationship**: Higher age correlates with higher income (men)
3. **Group Characteristics**
   - Men: older, moderate education, higher income
   - Women: younger, higher education, lower income

## Conclusion

Parallel Coordinate Plots are excellent for:

- Visualizing multivariate relationships
- Identifying patterns and clusters
- Exploring high-dimensional data
- Interactive data exploration

The key advantage: seeing **multiple dimensions** simultaneously, making hidden patterns visible.
