---
title:  "From Clutter to Clarity: Improving an SDG Time Series Chart"
subtitle: " Astep-by-step demo"
author: "Christophe Bontemps (SIAP)"
format:
  html:
    toc: true
    code-fold: false
  pdf:
    documentclass: article
    toc: true
execute:
  echo: false
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(gapminder)
library(plotly)

df_full <- gapminder

```


# Step 1: The Typical Excel Chart

Everything at once.

Problems:

- Too many colours
- Markers on every point
- Heavy gridlines
- Large legend
- Visual overload

```{r}

df <- df_full %>%
  filter(continent =="Asia" | continent == "Africa")

df%>%
ggplot() +
 aes(year,
     lifeExp,
     colour = country,
     group = country) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Life Expectancy by Country",
    x = "Year",
    y = "Life expectancy",
    colour = "Country"  ) +
 theme(
  legend.position = "bottom",
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.height = unit(0.2, "cm"),
  legend.key.width = unit(0.4, "cm")  ) +
  guides(
  colour = guide_legend(nrow = 6)
)
  
  
```



# Step 2: Remove Point Markers

Lines already show the trend.

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           colour = country,
           group = country)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Life Expectancy by Country",
    x = "Year",
    y = "Life expectancy",
    colour = "Country"
  ) +
 theme(
  legend.position = "bottom",
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.height = unit(0.2, "cm"),
  legend.key.width = unit(0.4, "cm")  ) +
  guides(
  colour = guide_legend(nrow = 3)
)
```



# Step 3: Reduce Visual Noise

Lighter gridlines and cleaner theme.

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           colour = country,
           group = country)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  )+
 theme(
  legend.position = "bottom",
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.height = unit(0.2, "cm"),
  legend.key.width = unit(0.4, "cm")  ) +
  guides(
  colour = guide_legend(nrow = 3)
)
```



# Step 4: Remove the Rainbow

Colour should communicate meaning.

Using one colour for everything.

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           group = country)) +
  geom_line(
    colour = "steelblue",
    linewidth = 0.7
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  )
```



# Step 5: Remove Unnecessary Borders

Focus attention on the data.

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           group = country)) +
  geom_line(
    colour = "steelblue",
    linewidth = 0.6
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  )
```



# Step 6: Highlight Regions Instead of Countries

Color now carries information.

```{r}
df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```


## Also using "small multiple" graphics

```{r}
df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  facet_wrap(vars(continent)) +
  labs(
    title = "Life Expectancy by continent",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
     legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```

## No legend needed
```{r}
df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  facet_wrap(vars(continent)) +
  labs(
    title = "Life Expectancy by continent",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
     legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```

```{r interactive}
p <- df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  facet_wrap(vars(continent)) +
  labs(
    title = "Life Expectancy by continent",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
     legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 

if (knitr::is_html_output()) {

  plotly::ggplotly(p)

} else {

  p
}
```


# Step 7: Ask a Question

How has a specific country evolved compared with all other countries?

```{r}

focus_country <- "Rwanda"

df <- df |>
  mutate(
    highlight =
      ifelse(country == focus_country, focus_country,  "Other countries")
  )

ggplot(df,
       aes(year,
           lifeExp,
           group = country,
           colour = highlight)) +
  geom_line(linewidth = 0.8) +
  theme_minimal(base_size = 14)
```



# Step 8: "*Le gris est mon ami*" (Grey is my friend )

Reference lines should stay in the background.

```{r}
ggplot() +

  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey70",
    alpha = .4,
    linewidth = .6
  ) +

  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = paste0(focus_country," Compared with Other Countries"),
    subtitle = "All countries shown as contextual references",
    x = NULL,
    y = "Life expectancy"
  )
```



# Step 9: Direct Labelling

The chart should work without a legend.

```{r}
focus_last <-
  df |>
  filter(country == focus_country) |>
  slice_max(year)

ggplot() +
  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey75",
    alpha = .35,
    linewidth = .5
  ) +
  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

  geom_text(
    data = focus_last,
    aes(year,
        lifeExp,
        label = country),
    hjust = 0.5,
    vjust = -1, 
    colour = "#08306B",
    fontface = "bold"
  ) +

  scale_x_continuous(
    limits = c(1952, 2010)
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  ) +

  labs(
    title = paste0(focus_country," Compared with Other Countries"),
    subtitle = "Other countries provide context",
    x = NULL,
    y = "Life expectancy"
  )
```



# Final Version*S*: Tell a story! 

Titles and annotations should guide the reader and tell stories!  

```{r}
focus_last <-
  df |>
  filter(country == focus_country) |>
  slice_max(year)

ggplot() +

  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey80",
    alpha = .30,
    linewidth = .5
  ) +

  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

   geom_text(
    data = focus_last,
    aes(year,
        lifeExp,
        label = country),
    hjust = 0.5,
    vjust = -1, 
    colour = "#08306B",
    fontface = "bold"
  ) +

  scale_x_continuous(
    limits = c(1952, 2010)
  ) +

  theme_minimal(base_size = 16) +

  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  ) +

  labs(
    title = paste0( focus_country,"'s Life Expectancy is back to pre-war levels"),
    subtitle = "Compared with all countries in the dataset",
    x = NULL,
    y = "Years"
  )
```


# Using analytics to tell other stories

> Ideas (TBC)
 - Highligt countries that have experienced a decrease
 - Use joyplot to highlight Africa vs Asia dynamics
 




# In conclusion

Design principles:


- Remove unnecessary ink
- Too many colors (Color is not decoration)
- Simplify background
- Identify a question
- Context vs focus
- Direct labelling
- Tell the story, not the dataset!




