---
title:  "From Clutter to Clarity: Improving a Time Series Chart"
subtitle: " A step-by-step demo"
author: "Christophe Bontemps (SIAP)"
format:
  revealjs:
    theme: solarized 
    css: myReveal.css
    slide-number: true
    transition: fade
    center: false
    progress: true
  html: 
    toc: true
    code-fold: true
  pdf:
    documentclass: article
    toc: true
execute:
  echo: false
  warning: false
  message: false
---



```{r}
library(tidyverse)
library(gapminder)
library(ggrepel)
library(plotly)
library(kableExtra)

df_full <- gapminder

```


# The Typical Excel Chart

```{r}
df <- df_full %>%
  filter(continent =="Asia" | continent == "Africa")

pExcel <-  df%>%
ggplot() +
 aes(year,
     lifeExp,
     colour = country,
     group = country) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Life Expectancy by Country",
    x = "Year",
    y = "Life expectancy",
    colour = "Country"  ) +
 theme(
  legend.position = "bottom",
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.height = unit(0.2, "cm"),
  legend.key.width = unit(0.4, "cm")  ) +
  guides(
  colour = guide_legend(nrow = 6)
)
pExcel   
  
```
## Problems:

- Too many colours
- Markers on every point
- Heavy gridlines
- Large legend
- Visual overload

```{r}
pExcel
```


# Remove to improve

Remove points to focus on lines (trends).

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           colour = country,
           group = country)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Life Expectancy by Country",
    x = "Year",
    y = "Life expectancy",
    colour = "Country"
  ) +
 theme(
  legend.position = "bottom",
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.height = unit(0.2, "cm"),
  legend.key.width = unit(0.4, "cm")  ) +
  guides(
  colour = guide_legend(nrow = 3)
)
```



## Reduce Visual Noise

Lighter gridlines and cleaner theme.

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           colour = country,
           group = country)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  )+
 theme(
  legend.position = "bottom",
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.height = unit(0.2, "cm"),
  legend.key.width = unit(0.4, "cm")  ) +
  guides(
  colour = guide_legend(nrow = 3)
)
```



# Remove the Rainbow

Color should communicate meaning, or help...


```{r}
ggplot(df,
       aes(year,
           lifeExp,
           group = country)) +
  geom_line(
    colour = "steelblue",
    linewidth = 0.7
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  )
```



# Remove Borders & Grids

Focus attention on the data.

```{r}
ggplot(df,
       aes(year,
           lifeExp,
           group = country)) +
  geom_line(
    colour = "steelblue",
    linewidth = 0.6
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  )
```



# Highlight Regions not Countries

Color now carries information.

```{r}
df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8, alpha = 1) +
  labs(
    title = "Life Expectancy by Country",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom", 
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```


## Also using "small multiple" graphics

Less lines is more information

```{r}
df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  facet_wrap(vars(continent)) +
  labs(
    title = "Life Expectancy by continent",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
     legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```

## Remove legend (not needed)

Avoid redundancy 

```{r}
df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  facet_wrap(vars(continent)) +
  labs(
    title = "Life Expectancy by continent",
    x = NULL,
    y = "Life expectancy"
  ) +
   theme_minimal(base_size = 14) +
  theme(
     legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```

## Use interactivity 

Helps identifying, not delivering a message

```{r interactive}
p <- df %>%
ggplot() +
       aes(year,
           lifeExp,
           group = country,
           colour = continent) +
  geom_line(linewidth = 0.8) +
  facet_wrap(vars(continent)) +
  labs(
    title = "Life Expectancy by continent",
    subtitle = "Interactive version",
    x = NULL,
    y = "Life expectancy"
  ) +
  theme_minimal(base_size = 14) +
  theme(
     legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 

if (knitr::is_html_output()) {

  plotly::ggplotly(p)

} else {

  p
}
```


# Ask a Question

```{r}
focus_country <- "Rwanda"
```


> How has a specific country (*`r focus_country`*) evolved?

```{r}

df <- df |>
  mutate(
    highlight =
      ifelse(country == focus_country, focus_country,  "Other countries")
  )

ggplot(df,
       aes(year,
           lifeExp,
           group = country,
           colour = highlight)) +
  geom_line(linewidth = 0.8) +
  theme_minimal(base_size = 14) +
  theme(
     legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
```
 

## Identify patterns

"*Le gris est mon ami*": Grey other lines as references

```{r}
ggplot() +
  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey50",
    alpha = 1,
    linewidth = .6
  ) +

  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = paste0(focus_country," Compared with Other Countries"),
    subtitle = "All countries shown as contextual references",
    x = NULL,
    y = "Life expectancy"
  )
```
## Use transparency

Overlapping transparent lines reveal patterns 


```{r}
ggplot() +
  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey50",
    alpha = .4,
    linewidth = .6
  ) +

  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = paste0(focus_country," Compared with Other Countries"),
    subtitle = "All countries shown as contextual references",
    x = NULL,
    y = "Life expectancy"
  )
```


## Direct Labelling

Provide labels close to information

```{r}
focus_last <-
  df |>
  filter(country == focus_country) |>
  slice_max(year)

ggplot() +
  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey50",
    alpha = .4,
    linewidth = .6
  ) +
  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

  geom_text(
    data = focus_last,
    aes(year,
        lifeExp,
        label = country),
    hjust = 0.5,
    vjust = -1, 
    colour = "#08306B",
    fontface = "bold"
  ) +

  scale_x_continuous(
    limits = c(1952, 2010)
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  ) +

  labs(
    title = paste0(focus_country," Compared with Other Countries"),
    subtitle = "Other countries provide context",
    x = NULL,
    y = "Life expectancy"
  )
```


##  Refine and annotate  

Titles and annotations should guide the reader and tell stories!  

```{r}
focus_last <-
  df |>
  filter(country == focus_country) |>
  slice_max(year)

p_focus <- ggplot() +
  geom_line(
    data = filter(df,
                  country != focus_country),
    aes(year,
        lifeExp,
        group = country),
    colour = "grey50",
    alpha = .4,
    linewidth = .6
  ) +

  geom_line(
    data = filter(df,
                  country == focus_country),
    aes(year,
        lifeExp),
    colour = "#08306B",
    linewidth = 1.2
  ) +

   geom_text(
    data = focus_last,
    aes(year,
        lifeExp,
        label = country),
    hjust = 0.5,
    vjust = -1, 
    colour = "#08306B",
    fontface = "bold"
  ) +

  scale_x_continuous(
    limits = c(1952, 2010)
  ) +

  theme_minimal(base_size = 16) +

  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  ) +

  labs(
    title = paste0(focus_country," Compared with other Countries"),
    #title = paste0( focus_country,"'s Life Expectancy is back to pre-war levels"),
    subtitle = "Compared with all countries in the dataset",
    x = NULL,
    y = "Years"
  )
p_focus
```

## Use anotations

Context always help the reader

```{r}
p_annotate <- p_focus + 
  annotate("rect", xmin = 1990, xmax = 1994,
                  ymin = 20, ymax = 30,
                  fill = "darkblue" , alpha = .1)+
                  
  annotate("text", x = 1992 , y = 30,
         color= "darkblue",
         alpha = 0.7,
         label = "Civil war")
p_annotate
  
```
## Provide a useful title
And mention source

```{r}
p_annotate  + 
  labs(
    title = paste0( focus_country,"'s Life Expectancy is back to pre-war levels"),
    subtitle = "A clear upside trend, but Life Expectancy is still quite low",
    caption= " Source: Gapminder (2026)",
    x = NULL,
    y = "Years"
  )
  
```



# Use analytics to detect stories

Detecting trends & highlighting countries 
 

```{r}
# Identifying countries with  decreases

Shocks <- df %>%
  arrange(country, year) %>%
  group_by(country, continent) %>%
  mutate(year_lag = lag(year),
         life_lag = lag(lifeExp),
         delta_years = year - year_lag,
         delta_life  = lifeExp - life_lag) %>%
  ungroup() %>%
  filter(
    delta_years >= 2,      # at least 2-year gap
    delta_years <= 5,      # at most 5-year gap
    delta_life <= -2       # drop of at least 2 years
  ) %>%
  arrange(country, continent, year, lifeExp, delta_years, delta_life)

# Shocks
```


```{r}
shock_countries <- Shocks %>%
  distinct(country) %>%
  pull(country)
# shock_countries

```


```{r}
df <-df %>%
  mutate(shock = country %in% shock_countries)
```



```{r}
shock_events <- df %>%
  arrange(country, year) %>%
  group_by(country, continent) %>%
  mutate(
    prev_year = lag(year),
    prev_life = lag(lifeExp),
    delta_year = year - prev_year,
    delta_life = lifeExp - prev_life
  ) %>%
  ungroup() %>%
  filter(
    delta_year >= 2,
    delta_year <= 5,
    delta_life <= -2
  ) %>%
  mutate(
    event_label = paste0(country, " (", prev_year, "→", year, ")")
  )

# kable(shock_events)
```



# Identifying shocks

```{r}
# Top N shocks
NbShocks <- 8

top_shocks <- shock_events %>%
  arrange(delta_life) %>%
  slice_head(n = NbShocks)

top_shocks %>%
  select(country, event_label,  prev_year,  prev_life, year, lifeExp, delta_life) %>%
         kable()

```



```{r}

# Base plot: all trajectories in background
p <- ggplot(df,
            aes(year, lifeExp, group = country)) +
  geom_line(
    colour = "grey85",
    alpha = 0.4,
    linewidth = 0.5
  )

## Adding segments with decreasing LE
p <- p +
  geom_segment(
    data = top_shocks,
    aes(
      x = prev_year,
      xend = year,
      y = prev_life,
      yend = lifeExp,
      group = country
    ),
    colour = "#B22222",
    linewidth = 0.5
  ) +
   geom_text_repel(
  data = top_shocks,
  aes(x = prev_year, y = prev_life, label = country),
  color = "#B22222",
  size = 3,
  box.padding = 0.4,
  point.padding = 0.3,
  direction = "both", 
  # segment.color = "#B22222",
  # segment.size = 0.5, 
  max.overlaps = Inf, 
  force = 1
) +
  labs(
    title = "Life Expectancy Shock Events",
    subtitle = "Highlighted segments with 2–5 year drops",
    x = NULL,
    y = "Life expectancy"
  )+
 theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) 
p




```

# A better story? 

Now that we have a story, on different continents

```{r}
shock_countries <- shock_events %>%
  arrange(delta_life) %>%
  slice_head(n = NbShocks) %>%
  distinct(country) %>%
  pull(country)
```

```{r}

ggplot() +

  # 1. All countries (background)
  geom_line(
    data = df,
    aes(year, lifeExp, group = country),
    colour = "grey85",
    alpha = 0.5,
    linewidth = 0.5
  ) +

  # 2. Highlight FULL trajectories of shock countries
  geom_line(
    data = df %>% 
      filter(country %in% shock_countries),
    aes(year, lifeExp, group = country, colour = continent),
    #colour = "darkgreen",
    linewidth = 0.5
  ) +

  # 3. Shock segments (event-level emphasis)
  geom_segment(
    data = top_shocks,
    aes(
      x = prev_year,
      xend = year,
      y = prev_life,
      yend = lifeExp
    ),
    colour = "#B22222",
    linewidth = 0.8
  ) +
  # 4. Direct label
  geom_text_repel(
  data = top_shocks,
  aes(x = prev_year, y = prev_life, label = country),
  color = "#B22222",
  size = 3,
  box.padding = 0.4,
  point.padding = 0.3,
  direction = "both", 
  # segment.color = "#B22222",
  # segment.size = 0.5, 
  max.overlaps = Inf, 
  force = 1
) +

  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +

  labs(
    title = "Life Expectancy Shock Events",
    subtitle = "Full trajectories of affected countries are highlighted",
    caption= " Source: Gapminder (2026)",
    x = NULL,
    y = "Life expectancy"
  )
```

# Conclusion

Design principles:


- Remove unnecessary ink
- Too many colors (Color is not decoration)
- Simplify background
- Identify a question
- Context vs focus
- Direct labelling
- Tell the story, not the dataset!




