---
title: "Who Holds the Seats? Women in Local Government, Worldwide and in the Pacific"
subtitle: "A scrollytelling exercise on classification choices in choropleth mapping"
format:
  closeread-html:
    remove-header-space: true
    cr-section:
      layout: overlay-center
    cr-style:
      narrative-background-color-overlay: "rgba(255,255,255,0.90)"
      narrative-text-color-overlay: "#1a1a1a"
      narrative-overlay-max-width: 460px
      narrative-border-radius: 10px
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)

# Read the data prepared by R/00_prepare_data.R. Run that script first if
# this file doesn't exist yet.
prepped <- readRDS("data/local_gov_women_clean.rds")

map_data     <- prepped$map_data
pacific_iso3 <- prepped$pacific_iso3

# Shared color scale -------------------------------------------------------
# Diverging around 30%, a commonly cited benchmark for the "critical mass"
# threshold in gender and representation literature. Adjust `threshold`
# below if you'd like to teach against a different reference point (e.g.
# 50% for full parity).
#
# The map (a fill) and the regional chart (a color) both use this same
# palette, so we define the shared parameters once and build both scales
# from them — that way a color change never has to be made twice.

threshold <- 30

diverging_palette <- list(
  low      = "#B2182B",
  mid      = "#F7F7F7",
  high     = "#2166AC",
  midpoint = threshold,
  limits   = c(0, 100)
)

fill_scale <- do.call(scale_fill_gradient2, c(diverging_palette, list(
  name     = "Seats held\nby women (%)",
  na.value = "grey85",
  labels   = label_number(suffix = "%")
)))

color_scale <- do.call(scale_color_gradient2, c(diverging_palette, list(
  name     = "Seats held\nby women (%)",
  na.value = "grey60",
  labels   = label_number(suffix = "%")
)))

map_theme <- theme_void(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.caption    = element_text(size = 9, color = "grey40", hjust = 0)
  )

n_reported <- sum(map_data$status == "reported", na.rm = TRUE)

# ---- Regional summary data ------------------------------------------------

regional_data <- map_data %>%
  sf::st_drop_geometry() %>%
  filter(status == "reported", !is.na(subregion))

subregion_order <- regional_data %>%
  group_by(subregion) %>%
  summarize(median_value = median(value, na.rm = TRUE)) %>%
  arrange(median_value) %>%
  pull(subregion)

regional_data <- regional_data %>%
  mutate(subregion = factor(subregion, levels = subregion_order))

# Which subregions count as "Asia-Pacific" for the focus window on the
# regional chart. This follows the broad ESCAP sense of the term (Asia
# plus the Pacific) rather than just the Pacific islands. Adjust this list
# if your course defines the region differently (e.g. whether to include
# Western Asia, or Australia and New Zealand).
asia_pacific_subregions <- c(
  "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia",
  "Melanesia", "Micronesia", "Polynesia", "Australia and New Zealand"
)

# ---- Labels baked into the single map for the Pacific zoom ----------------
# These are small at the global view and only become legible once the
# pan/zoom focus effects below scale the map up around the Pacific.
#
# Every country here sits between roughly 140 and 180 degrees E, so we
# stay on the standard -180/180 map and never need to touch the
# antimeridian at all.

lowest_iso3    <- c("FSM", "KIR", "VUT", "TUV", "MHL")
highlight_iso3 <- c(lowest_iso3, "FJI", "NCL")

label_points <- map_data %>%
  filter(iso3 %in% highlight_iso3, status == "reported") %>%
  # st_point_on_surface (rather than st_centroid) guarantees a point that
  # actually falls on the country's own polygon. This matters here because
  # Kiribati's territory is split across three island groups spanning
  # roughly -150 to 173 degrees — a plain centroid could land in open
  # ocean, or even outside the Pacific crop entirely.
  mutate(anchor = sf::st_point_on_surface(geometry)) %>%
  mutate(
    lon = sf::st_coordinates(anchor)[, 1],
    lat = sf::st_coordinates(anchor)[, 2]
  ) %>%
  sf::st_drop_geometry() %>%
  mutate(is_lowest = iso3 %in% lowest_iso3)

# Kiribati specifically: its largest island by area, Kiritimati, is out in
# the Line Islands (~-157), far from the Gilbert Islands group (~173 E)
# where the capital, Tarawa, sits. Anchor its label at Tarawa by hand so
# it lands inside the Pacific crop instead of wherever st_point_on_surface
# happens to pick.
label_points <- label_points %>%
  mutate(
    lon = ifelse(iso3 == "KIR", 173.0, lon),
    lat = ifelse(iso3 == "KIR", 1.4, lat)
  )

# ---- Map pan/zoom targets --------------------------------------------------
# coord_sf below is fixed at these limits for every version of the map, so
# the percentages calculated here stay valid across triggers. This is the
# plain, standard -180/180 world map.
world_xlim <- c(-180, 180)
world_ylim <- c(-60, 85)
world_cx   <- mean(world_xlim)
world_cy   <- mean(world_ylim)

# A small helper so every zoom target is calculated the same way, instead
# of repeating this arithmetic for Africa, Asia-Pacific, and the Pacific.
map_pan_pct <- function(cx, cy) {
  list(
    x = round(100 * (world_cx - cx) / diff(world_xlim), 1),
    y = round(100 * (world_cy - cy) / diff(world_ylim), 1)
  )
}

africa_pan       <- map_pan_pct(cx = 17,  cy = 2)   # centered roughly on Central/East Africa
asia_pacific_pan <- map_pan_pct(cx = 115, cy = 15)  # wide shot: South Asia through the western Pacific
pacific_pan      <- map_pan_pct(cx = 162, cy = -8)  # tight shot: the Pacific island countries

# ---- Regional chart pan/zoom target ----------------------------------------
# The chart's y-axis is the subregion factor, plotted at integer positions
# 1..n in the order of subregion_order (lowest median at the bottom). We
# find where the Asia-Pacific subregions fall in that ordering so the
# focus window lands on the right band of rows, whatever that order turns
# out to be once real data is plotted.
n_subregions      <- length(subregion_order)
asia_pacific_ranks <- which(subregion_order %in% asia_pacific_subregions)
regional_focus_cy  <- mean(range(asia_pacific_ranks))
regional_focus_span <- diff(range(asia_pacific_ranks)) + 1

regional_pan_y_pct  <- round(100 * ((n_subregions + 1) / 2 - regional_focus_cy) / n_subregions, 1)
regional_scale_by   <- round(n_subregions / regional_focus_span, 1)
```

Local government is often the level closest to people's daily lives — and
in most countries, women remain a minority of the elected officials who
run it. SDG indicator 5.5.1b tracks the share of seats held by women in
local deliberative bodies. This page walks through what that indicator
looks like globally, then zooms first into Africa and then into Asia and
the Pacific, before settling on the Pacific island countries, where
several of the lowest values in the dataset are found.

::: {.callout-note}
**For the classification-methods lesson:** as you scroll, notice how the
same underlying numbers can look quite different depending on where you
place the color break (here, at 30%). Try changing `threshold` in the
setup chunk to 50% and re-rendering to see how the map's story shifts.
:::

:::{.cr-section}

Across the `r n_reported` countries with reported data, the share of
women in local government ranges from under 2% to over 50%. @cr-map

:::{#cr-map}
```{r}
#| label: world-map
#| fig-width: 10
#| fig-height: 6.5
ggplot(map_data) +
  geom_sf(aes(fill = value), color = "white", linewidth = 0.08) +
  geom_text(
    data = label_points,
    aes(x = lon, y = lat, label = admin, fontface = ifelse(is_lowest, "bold", "plain")),
    size = 2.1, color = "grey15", check_overlap = TRUE
  ) +
  # datum = NA turns off coord_sf's automatic lat/long graticule — good
  # hygiene generally, and rules out graticule breaks as a source of any
  # stray lines.
  coord_sf(xlim = world_xlim, ylim = world_ylim, expand = FALSE, datum = NA) +
  fill_scale +
  map_theme
```
:::

The world map uses a **diverging classification centered on
`r threshold`%**: countries below that line shade toward red, countries
above shade toward blue. Grey means no elected local government body, or
no data reported. This single choice already does a lot of storytelling
— a different center would draw the eye somewhere else entirely.

No single global pattern jumps out immediately — high and low values are
scattered across regions rather than clustered in one place. @cr-regional

:::{#cr-regional}
```{r}
#| label: regional-chart
#| fig-width: 8
#| fig-height: 7.5
ggplot(regional_data, aes(x = value, y = subregion)) +
  geom_vline(xintercept = threshold, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = value), size = 2.6, alpha = 0.9) +
  stat_summary(
    fun = median, geom = "point", shape = 18, size = 4, color = "grey20"
  ) +
  color_scale +
  scale_x_continuous(labels = label_number(suffix = "%"), limits = c(0, 100)) +
  labs(x = "Seats held by women (%)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.caption    = element_text(size = 9, color = "grey40", hjust = 0)
  )
```
:::

Ordering subregions by their median value (grey diamonds), and coloring
each country the same way as the map, shows the same red-to-blue spread
playing out region by region rather than country by country.

Let's put a window around Asia and the Pacific specifically — the region
this course is built around.
[@cr-regional]{pan-to="0%,`r regional_pan_y_pct`%" scale-by="`r regional_scale_by`"}

That band covers everywhere from Central Asia to the Pacific islands, and
it already hints at how wide the spread within the region is — we'll see
exactly how wide once we get to the map. [@cr-regional]{pan-to="0%,`r regional_pan_y_pct`%" scale-by="`r regional_scale_by`"}

Let's go back to the map, starting with a different region first.
@cr-map

Zooming into Africa: reported values here span almost the entire range on
their own. Benin (3.5%) and Ghana (3.8%) rank among the lowest anywhere
in the dataset, while Rwanda (43.6%), Namibia (44.9%), Senegal (47.2%),
and Tunisia (48.5%) sit close to parity. Eight African countries in this
dataset — including Egypt, Somalia, and Sudan — report no elected local
government body at all, shown here in grey.
[@cr-map]{pan-to="`r africa_pan$x`%,`r africa_pan$y`%" scale-by="2.6"}

Now let's turn to Asia and the Pacific together — not just the Pacific
islands, but the wider region.
[@cr-map]{pan-to="`r asia_pacific_pan$x`%,`r asia_pacific_pan$y`%" scale-by="1.7"}

This region holds both ends of the entire dataset. Oman (0.8%) and Saudi
Arabia (1.2%) report the two lowest shares recorded anywhere, while India
(44.4%) and Nepal (41.3%) sit close to parity — both countries legally
reserve a share of local council seats for women, a policy choice with a
visible effect on the map.
[@cr-map]{pan-to="`r asia_pacific_pan$x`%,`r asia_pacific_pan$y`%" scale-by="1.7"}

Let's zoom in a second time, further into the Pacific specifically.
[@cr-map]{pan-to="`r pacific_pan$x`%,`r pacific_pan$y`%" scale-by="3.2"}

Reported values across the Pacific island countries and territories in
this dataset span from about 1.5% to 47%. Samoa, Tonga, and American
Samoa report no elected local government body at all, so they're greyed
out rather than colored. [@cr-map]{pan-to="`r pacific_pan$x`%,`r pacific_pan$y`%" scale-by="4.5"}

Micronesia (1.5%), Kiribati (7.3%), Vanuatu (10.1%), Tuvalu (12.5%), and
the Marshall Islands (15.9%) report some of the lowest shares in the
entire dataset — now legible as the map zooms in. Fiji (35.8%) and New
Caledonia (47.2%) sit at the other end of the regional range, a reminder
that "the Pacific" is not one story, but many.
[@cr-map]{pan-to="`r pacific_pan$x`%,`r pacific_pan$y`%" scale-by="5.5"}

Zooming back out: this pattern is one piece of a much bigger picture.
@cr-map

:::

## Discussion prompt

For the Moodle exercise: ask participants to identify one country whose
color would change categories if the classification break moved from 30%
to 50%. Then ask what argument they would make for using one threshold
over the other in a report aimed at policymakers versus one aimed at
the general public.

::: {.callout-tip}
**Rendering this document:** run `R/00_prepare_data.R` once to produce
`data/local_gov_women_clean.rds`, then render `index.qmd` as usual
(`quarto render index.qmd`). You'll need the closeread extension
installed in this directory (`quarto add qmd-lab/closeread`) and the R
packages listed at the top of the setup chunk, plus `rnaturalearth`,
`rnaturalearthdata`, and `countrycode` for the data prep script.
:::

::: {.callout-important}
**About the pan-to/scale-by numbers:** all of them — `africa_pan`,
`asia_pacific_pan`, `pacific_pan` for the map, and `regional_pan_y_pct`/
`regional_scale_by` for the chart — are calculated in the setup chunk
from approximate region centers, not verified against a real render.
Use `quarto preview index.qmd` (live reload) and nudge the `cx`/`cy`
values passed to `map_pan_pct()` and the `scale-by` numbers in the
narrative triggers until each zoom lands where you want it. Because
they're computed rather than hard-coded, adjusting a center updates
every trigger that uses it automatically.
:::
