---
title: "Who Holds the Seats? Women in Local Government"
subtitle: "A scrollytelling exercise"

format:
  closeread-html:
    remove-header-space: true
    cr-section:
      layout: overlay-center
    cr-style:
      narrative-background-color-overlay: "rgba(255,255,255,0.90)"
      narrative-text-color-overlay: "#1a1a1a"
      narrative-overlay-max-width: 460px
      narrative-border-radius: 10px
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r, setup}
library(sf)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)

# Run 00_prepare_data.R first if
# this file doesn't exist yet.

prepped <- readRDS("data/local_gov_women_clean.rds")

map_data     <- prepped$map_data
pacific_iso3 <- prepped$pacific_iso3

# Shared color scale -------------------------------------------------------
# Diverging threshold in gender and representation l

threshold <- 30

diverging_palette <- list(
  low      = "#B2182B",
  mid      = "#F7F7F7",
  high     = "#2166AC",
  midpoint = threshold,
  limits   = c(0, 100)
)

fill_scale <- do.call(scale_fill_gradient2, c(diverging_palette, list(
  name     = "Seats held\nby women (%)",
  na.value = "grey85",
  labels   = label_number(suffix = "%")
)))

color_scale <- do.call(scale_color_gradient2, c(diverging_palette, list(
  name     = "Seats held\nby women (%)",
  na.value = "grey60",
  labels   = label_number(suffix = "%")
)))

map_theme <- theme_void(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.caption    = element_text(size = 9, color = "grey40", hjust = 0)
  )

n_reported <- sum(map_data$status == "reported", na.rm = TRUE)

# ---- Regional summary data ------------------------------------------------

regional_data <- map_data %>%
  sf::st_drop_geometry() %>%
  filter(status == "reported", !is.na(subregion))

subregion_order <- regional_data %>%
  group_by(subregion) %>%
  summarize(median_value = median(value, na.rm = TRUE)) %>%
  arrange(median_value) %>%
  pull(subregion)

regional_data <- regional_data %>%
  mutate(subregion = factor(subregion, levels = subregion_order))


# ---- Samll Islands   for the Pacific zoom ----------------

lowest_iso3    <- c( "KIR", "VUT")
highlight_iso3 <- c(lowest_iso3, "FJI", "NCL")

label_points <- map_data %>%
  filter(iso3 %in% highlight_iso3, status == "reported") %>%
  # st_point_on_surface (rather than st_centroid) guarantees a point that actually falls on the country's own polygon. 
  mutate(anchor = sf::st_point_on_surface(geometry)) %>%
  mutate(
    lon = sf::st_coordinates(anchor)[, 1],
    lat = sf::st_coordinates(anchor)[, 2]
  ) %>%
  sf::st_drop_geometry() %>%
  mutate(is_lowest = iso3 %in% lowest_iso3)

# Kiribati specifics

label_points <- label_points %>%
  mutate(
    lon = ifelse(iso3 == "KIR", 173.0, lon),
    lat = ifelse(iso3 == "KIR", 1.4, lat)
  )

# ---- Map pan/zoom targets --------------------------------------------------
world_xlim <- c(-180, 180)
world_ylim <- c(-60, 85)
world_cx   <- mean(world_xlim)
world_cy   <- mean(world_ylim)

# A small helper so every zoom target is calculated the same way, instead
# of repeating this arithmetic for Africa, Asia-Pacific, and the Pacific.
# THIS MAY BE AN ISSUE....

map_pan_pct <- function(cx, cy) {
  list(
    x = round(100 * (world_cx - cx) / diff(world_xlim), 1),
    y = round(100 * (world_cy - cy) / diff(world_ylim), 1)
  )
}

africa_pan       <- map_pan_pct(cx = 17,  cy = 2)   # centered roughly on Central/East Africa
asia_pacific_pan <- map_pan_pct(cx = 200, cy = 15)  # wide shot: South Asia through the western Pacific
pacific_pan      <- map_pan_pct(cx = 300, cy = 2)  # tight shot: the Pacific island countries

# ---- Point chart pan/zoom target ----------------------------------------

n_subregions      <- length(subregion_order)
melanesia_rank <- which(subregion_order == "Melanesia")

regional_focus_cy   <- mean(c(1, melanesia_rank))
regional_focus_span <- melanesia_rank

regional_pan_y_pct  <- round(100 * ((n_subregions + 1) / 2 - regional_focus_cy) / n_subregions, 1)
regional_scale_by   <- round(n_subregions / regional_focus_span, 1)
```

# Scrollytelling with maps

*Christophe Bontemps (SIAP)*

*Scrollytelling* combines narrative text with interactive maps to reveal information progressively as the reader scrolls. By synchronizing map movements and visual changes with the story, it helps audiences explore complex geographic data in a clear and engaging way.

In this demo, we will see how map-based scrollytelling can guide readers through a complex choropleth map representing Women in local governments all over the world, highlighting key locations and insights step by step, region by region. 

> Just scroll down!


:::{.cr-section}

Across the `r n_reported` countries with reported data, the share of
women in local government ranges from under **2%** to over **50%**. @cr-map


:::{#cr-map}
```{r}
#| label: world-map
#| fig-width: 15
#| fig-height: 8.5
ggplot(map_data) +
  geom_sf(aes(fill = value), color = "white", linewidth = 0.08) +
  geom_text(
    data = label_points,
    aes(x = lon, y = lat+3, label = admin, fontface =  "bold"),
    size = 2.1, color = "grey15", check_overlap = TRUE
  ) +
  # datum = NA turns off coord_sf's automatic lat/long graticule — good
  # hygiene generally, and rules out graticule breaks as a source of any
  # stray lines.
  coord_sf(xlim = world_xlim, ylim = world_ylim, expand = FALSE) +
  fill_scale +
  map_theme
```
:::

The  map uses a **diverging classification palette** centered on
`r threshold`%: countries below shade toward red, countries
above shade toward blue. Grey means no elected local government body, or
no data reported.

This single choice already does a lot of storytelling:
*Different choices would draw the eye somewhere else*.


No single global pattern jumps out immediately — high and low values are
scattered across regions rather than clustered in one place. @cr-regional

Let's do something else. 

:::{#cr-regional}
```{r}
#| label: regional-chart
#| fig-width: 9
#| fig-height: 7.5
ggplot(regional_data, aes(x = value, y = subregion)) +
  #geom_vline(xintercept = threshold, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = value), size = 2.6, alpha = 0.99) +
  stat_summary(
    fun = median, geom = "point", shape = 3, size = 2, color = "grey60"
  ) +
  color_scale +
  scale_x_continuous(labels = label_number(suffix = "%"), limits = c(0, 100)) +
  labs(x = "Seats held by women (%)  ", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.caption    = element_text(size = 9, color = "grey40", hjust = 0)
  )
```
:::

Ordering **subregions** by their median value (cross), and coloring
each country with the same red-to-blue spread
provides a regional focus and **reveals some patterns**, while highlighting the **diversity** within subregions. @cr-regional

Let's focus on the upper part of the table specifically.
[@cr-regional]{pan-to="0%,50%" scale-by="1.5"}

Even in regions with a good proportion of women in local governmental bodies, the **diversity** of values is striking (*e.g.* Melanesia). 


Let's go back to the map and have a closer look by region. [@cr-regional]{pan-to="0%,`r regional_pan_y_pct`%" scale-by="`r regional_scale_by`"}

Let's start with **Africa** first.
@cr-map

Zooming into Africa: **Benin** (3.5%) and **Ghana** (3.8%) rank among the lowest anywhere
in the dataset, while **Rwanda** (43.6%), **Namibia** (44.9%) and **Senegal** (47.2%),
sit close to parity. [@cr-map]{pan-to="`r africa_pan$x`%,`r africa_pan$y`%" scale-by="2.6"}

Eight African countries in this
dataset — including Egypt, Somalia, and Sudan — report no elected local
government body at all, shown here in grey.


Now let's turn to **Asia and the Pacific**, home of ESCAP.
[@cr-map]{pan-to="`r asia_pacific_pan$x`%,`r asia_pacific_pan$y`%" scale-by="1.7"}

This region holds both ends of the entire dataset. **Oman** (0.8%) and **Saudi
Arabia** (1.2%) report the two lowest shares recorded, while **India**
(44.4%) and **Nepal** (41.3%) sit close to parity. A policy choice with a
visible effect on the map.


Let's zoom in a second time, further into **the Pacific** specifically.
[@cr-map]{pan-to="-150%, -2%" scale-by="4"}



**Micronesia** (1.5%), **Kiribati (7.3%)** and **Vanuatu** (10.1%) report some of the lowest shares in the entire dataset.

**Fiji** (35.8%) and **New Caledonia** (47.2%) sit at the other end of the regional range, a reminder
that "the Pacific" is not one story, but many.


Zooming back to the map highlights the difficulty of exploring this information only through a map where countries with large areas are largely over represented.
@cr-map

:::

# To conclude
Scrollytelling maps offer a powerful way to combine narrative, data, and geography into a single interactive experience. By synchronizing movements on a map (zooming, panning, highlighting) with explanatory text, they guide readers through complex spatial information in a clear and engaging way.

Effective scrollytelling is not simply about adding animation. Rather, it is about using animation and data visualization to support a compelling narrative, draw attention to key insights, and help the audience develop a deeper understanding of the information being presented.


