---
title: "Who Holds the Seats? Women in Local Government, Worldwide and in the Pacific"
subtitle: "A scrollytelling exercise on classification choices in choropleth mapping"
format:
  closeread-html:
    remove-header-space: true
    cr-section:
      layout: overlay-center
    cr-style:
      narrative-background-color-overlay: "rgba(255,255,255,0.90)"
      narrative-text-color-overlay: "#1a1a1a"
      narrative-overlay-max-width: 460px
      narrative-border-radius: 10px
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)

# Read the data prepared by R/00_prepare_data.R. Run that script first if
# this file doesn't exist yet.
prepped <- readRDS("data/local_gov_women_clean.rds")

# Read the data prepared by R/00_prepare_data.R. Run that script first if
# this file doesn't exist yet.
prepped <- readRDS("data/local_gov_women_clean.rds")

map_data     <- prepped$map_data
pacific_iso3 <- prepped$pacific_iso3

# Shared color scale -------------------------------------------------------
# Diverging around 30%, a commonly cited benchmark for the "critical mass"
# threshold in gender and representation literature. Adjust `threshold`
# below if you'd like to teach against a different reference point (e.g.
# 50% for full parity).

threshold <- 30

fill_scale <- scale_fill_gradient2(
  name     = "Seats held\nby women (%)",
  low      = "#B2182B",
  mid      = "#F7F7F7",
  high     = "#2166AC",
  midpoint = threshold,
  na.value = "grey85",
  limits   = c(0, 100),
  labels   = label_number(suffix = "%")
)

map_theme <- theme_void(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.caption    = element_text(size = 9, color = "grey40", hjust = 0)
  )

n_reported <- sum(map_data$status == "reported", na.rm = TRUE)

# ---- Regional summary data ------------------------------------------------

regional_data <- map_data %>%
  sf::st_drop_geometry() %>%
  filter(status == "reported", !is.na(subregion)) %>%
  mutate(
    is_pacific_subregion = subregion %in% c("Melanesia", "Micronesia", "Polynesia")
  )

subregion_order <- regional_data %>%
  group_by(subregion) %>%
  summarize(median_value = median(value, na.rm = TRUE)) %>%
  arrange(median_value) %>%
  pull(subregion)

regional_data <- regional_data %>%
  mutate(subregion = factor(subregion, levels = subregion_order))

# ---- Labels baked into the single map for the Pacific zoom ----------------
# These are small at the global view and only become legible once the
# pan/zoom focus effects below scale the map up around the Pacific.
#
# Every country here sits between roughly 140 and 180 degrees E, so we
# stay on the standard -180/180 map and never need to touch the
# antimeridian at all.

lowest_iso3    <- c("FSM", "KIR", "VUT", "TUV", "MHL")
highlight_iso3 <- c(lowest_iso3, "FJI", "NCL")

label_points <- map_data %>%
  filter(iso3 %in% highlight_iso3, status == "reported") %>%
  # st_point_on_surface (rather than st_centroid) guarantees a point that
  # actually falls on the country's own polygon. This matters here because
  # Kiribati's territory is split across three island groups spanning
  # roughly -150 to 173 degrees — a plain centroid could land in open
  # ocean, or even outside the Pacific crop entirely.
  mutate(anchor = sf::st_point_on_surface(geometry)) %>%
  mutate(
    lon = sf::st_coordinates(anchor)[, 1],
    lat = sf::st_coordinates(anchor)[, 2]
  ) %>%
  sf::st_drop_geometry() %>%
  mutate(is_lowest = iso3 %in% lowest_iso3)

# Kiribati specifically: its largest island by area, Kiritimati, is out in
# the Line Islands (~-157), far from the Gilbert Islands group (~173 E)
# where the capital, Tarawa, sits. Anchor its label at Tarawa by hand so
# it lands inside the Pacific crop instead of wherever st_point_on_surface
# happens to pick.
label_points <- label_points %>%
  mutate(
    lon = ifelse(iso3 == "KIR", 173.0, lon),
    lat = ifelse(iso3 == "KIR", 1.4, lat)
  )

# ---- Full-map extent, used to work out pan-to percentages -----------------
# coord_sf below is fixed at these limits for every version of the map, so
# the percentages calculated here stay valid across triggers. This is the
# plain, standard -180/180 world map.
world_xlim <- c(-180, 180)
world_ylim <- c(-60, 85)
world_cx   <- mean(world_xlim)
world_cy   <- mean(world_ylim)

# Target center for the Pacific zoom. Chosen to sit inside the cluster of
# countries we highlight (roughly 140-180 E), which stays on one side of
# the antimeridian on purpose.
pacific_cx <- 162
pacific_cy <- -8

pan_x_pct <- round(100 * (world_cx - pacific_cx) / diff(world_xlim), 1)
pan_y_pct <- round(100 * (world_cy - pacific_cy) / diff(world_ylim), 1)
```

Local government is often the level closest to people's daily lives — and
in most countries, women remain a minority of the elected officials who
run it. SDG indicator 5.5.1b tracks the share of seats held by women in
local deliberative bodies. This page walks through what that indicator
looks like globally, then zooms into the Pacific, where several of the
lowest values in the dataset are found.

::: {.callout-note}
**For the classification-methods lesson:** as you scroll, notice how the
same underlying numbers can look quite different depending on where you
place the color break (here, at 30%). Try changing `threshold` in the
setup chunk to 50% and re-rendering to see how the map's story shifts.
:::

:::{.cr-section}

Across the `r n_reported` countries with reported data, the share of
women in local government ranges from under 2% to over 50%. @cr-map

:::{#cr-map}
```{r}
#| label: world-map
#| fig-width: 10
#| fig-height: 6.5
ggplot(map_data) +
  geom_sf(aes(fill = value), color = "white", linewidth = 0.08) +
  geom_text(
    data = label_points,
    aes(x = lon, y = lat, label = admin, fontface = ifelse(is_lowest, "bold", "plain")),
    size = 2.1, color = "grey15", check_overlap = TRUE
  ) +
  # datum = NA turns off coord_sf's automatic lat/long graticule. Without
  # it, the graticule's default break algorithm can misbehave on the
  # unusual 0-360 xlim used here (needed for the Pacific-centered shift)
  # and render as long horizontal stripes across the map.
  coord_sf(xlim = world_xlim, ylim = world_ylim, expand = FALSE, datum = NA) +
  fill_scale +
  map_theme
```
:::

The world map uses a **diverging classification centered on
`r threshold`%**: countries below that line shade toward red, countries
above shade toward blue. Grey means no elected local government body, or
no data reported. This single choice already does a lot of storytelling
— a different center would draw the eye somewhere else entirely.

No single global pattern jumps out immediately — high and low values are
scattered across regions rather than clustered in one place. @cr-regional

:::{#cr-regional}
```{r}
#| label: regional-chart
#| fig-width: 8
#| fig-height: 7.5
ggplot(regional_data, aes(x = value, y = subregion)) +
  geom_vline(xintercept = threshold, linetype = "dashed", color = "grey50") +
  geom_point(
    aes(color = is_pacific_subregion),
    size = 2.6, alpha = 0.85
  ) +
  stat_summary(
    fun = median, geom = "point", shape = 18, size = 4, color = "black"
  ) +
  scale_color_manual(
    values = c(`TRUE` = "#B2182B", `FALSE` = "grey55"),
    guide  = "none"
  ) +
  scale_x_continuous(labels = label_number(suffix = "%"), limits = c(0, 100)) +
  labs(x = "Seats held by women (%)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.caption = element_text(size = 9, color = "grey40", hjust = 0))
```
:::

Ordering subregions by their median value (black diamonds) shows that
the Pacific groupings — Melanesia, Micronesia, and Polynesia, highlighted
in red — sit toward the lower end of the distribution.

Let's zoom in, right on the map itself. [@cr-map]{pan-to="`r pan_x_pct`%,`r pan_y_pct`%" scale-by="3.2"}

Reported values across the Pacific island countries and territories in
this dataset span from about 1.5% to 47%. Samoa, Tonga, and American
Samoa report no elected local government body at all, so they're greyed
out rather than colored. [@cr-map]{pan-to="`r pan_x_pct`%,`r pan_y_pct`%" scale-by="4.5"}

Micronesia (1.5%), Kiribati (7.3%), Vanuatu (10.1%), Tuvalu (12.5%), and
the Marshall Islands (15.9%) report some of the lowest shares in the
entire dataset — now legible as the map zooms in. Fiji (35.8%) and New
Caledonia (47.2%) sit at the other end of the regional range, a reminder
that "the Pacific" is not one story, but many.
[@cr-map]{pan-to="`r pan_x_pct`%,`r pan_y_pct`%" scale-by="5.5"}

Zooming back out: this pattern is one piece of a much bigger picture.
@cr-map

:::

## Discussion prompt

For the Moodle exercise: ask participants to identify one country whose
color would change categories if the classification break moved from 30%
to 50%. Then ask what argument they would make for using one threshold
over the other in a report aimed at policymakers versus one aimed at
the general public.

::: {.callout-tip}
**Rendering this document:** run `R/00_prepare_data.R` once to produce
`data/local_gov_women_clean.rds`, then render `index.qmd` as usual
(`quarto render index.qmd`). You'll need the closeread extension
installed in this directory (`quarto add qmd-lab/closeread`) and the R
packages listed at the top of the setup chunk, plus `rnaturalearth`,
`rnaturalearthdata`, and `countrycode` for the data prep script.
:::

::: {.callout-important}
**About the pan-to/scale-by numbers above:** I calculated `pan_x_pct` and
`pan_y_pct` from the map's plotted extent and the Pacific's approximate
center, but I haven't been able to render this document myself to check
them, since I don't have R or Quarto available where I built this. Use
`quarto preview index.qmd` (it live-reloads as you save) and expect to
nudge the `scale-by` values and the two percentages by eye — a few tries
is normal for this effect. If the zoom keeps landing off-center, the
likely culprit is `world_xlim`/`world_ylim` or `pacific_cx`/`pacific_cy`
in the setup chunk not quite matching what actually gets plotted; adjust
those and the percentages will update automatically since they're
computed, not hard-coded.
:::

::: {.callout-important}
**About the stripe artifacts you saw:** those came from trying to
recenter the whole world map around the Pacific (longitude-shifting so
it wouldn't be split at the edges). That approach is genuinely fiddly in
`sf` — countries whose polygons already cross the antimeridian in the
source data (Russia, Fiji) don't shift cleanly, and produce exactly that
kind of artifact. This draft no longer does that at all: it's back to
the plain, standard -180/180 world map, and the Pacific zoom is done
with a CSS pan/scale effect on a region (roughly 140-180 degrees E) that
never crosses the antimeridian in the first place. `coord_sf(..., datum = NA)`
is still in place too — that turns off `coord_sf`'s automatic lat/long
graticule, which is good practice regardless and rules out a second
possible source of stray lines.

One casualty of simplifying this way: Kiribati's Line Islands, far to
the east, won't appear on the zoomed Pacific view — only its main
Gilbert Islands group will, which is why Kiribati's label is anchored
manually near Tarawa (the capital) in the setup chunk rather than
computed automatically.
:::
:::
